### 2017-10-01 paper/sexSelectionSpacing-ver1.tex
Add spell 3 files and update paper

\title{Son Preference, Birth Spacing, and Sex-Selective Abortions%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to three anonymous referees, Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of Aarhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support from the University of Washington Royalty Research Fund and the 
Development Research Group of the World Bank is gratefully acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
% Prior versions of this paper were presented under the title ``The 
% Determinants of sex-selective Abortions.''
}
}

@@

\author{Claus C P\"ortner\\
    Department of Economics\\
    Albers School of Business and Economics\\
    Seattle University, P.O. Box 222000\\
    Seattle, WA 98122\\
    \href{mailto:cportner@seattleu.edu}{\texttt{cportner@seattleu.edu}}\\
    \href{http://www.clausportner.com}{\texttt{www.clausportner.com}}\\
    \& \\
    Center for Studies in Demography and Ecology \\
    University of Washington\\ \vspace{2cm}
    }

\date{April 2017}

@@

% Demography abstract
\noindent This paper examines how birth spacing interact with the use of 
sex-selective abortions. 
I introduce a statistical method that incorporates that sex-selective abortions 
affect both the likelihood of a son and spacing between births.
Using India's National Family and Health Surveys,
I show that use of sex selection leads to \emph{longer} spacing after a daughter 
is born than after a son.
Women with 8 or more years of education, both in urban and rural areas, are 
the main users of sex-selective abortions, whereas women with less education 
do not appear to use sex selection.
Predicted lifetime fertility for high-education women declined more than 10\% 
between 1985--1994, when sex selection was legal, and 1995--2006, when sex selection 
was illegal.
Abortions per woman increased almost 20\% for urban women and 50\%
for rural women between the two periods, suggesting that making sex 
selection illegal has not reversed its use.
Finally, sex selection appears to be used to ensure one son rather 
than multiple sons.

@@

% [EXPLAIN RESULTS HERE]
% Main findings!
There are three main findings.

There is, however, still no evidence of sex selection being used on the first birth.

Secondly, sex selection appears to be used for securing one son, rather than a large 
number of sons.
There is only limited use of sex-selective abortions for better-educated
women with one or more sons.
The exception is rural women with one son and one daughter, presumably to compensate for 
the higher child mortality in rural areas.
These results are in line with the differential stopping behavior observed in many studies
before sex selection became available \citep{repetto72,arnold98,dreze01}.

@@

[How to compare survival curves?]

@@

One solution to this problem is to condition on the parity progression
rates.
That is, instead of comparing say how long it takes for 50\% of women
who began the spell to have the next birth across boy and girl composition, 
which would work well if the next birth was nearly universal, we 
could calculate how long it takes for 50\% of those women who will 
eventually have a next child to have that child based on the estimated
survival curves.
This number can then be compared across sex compositions to 
understand if the spell length for girl composition increases
more than for boy composition with the introduction of sex selection.

Obviously, this should be calculated for different percentages.
Actually, it probably makes more sense to show this graphically
instead. 
All compositions would begin at 100\% and end at 0\% because
we condition on having the next child.

@@

% \begin{figure}[htp]%
% \centering
% \includegraphics[width=0.9\textwidth]{sex_ratio_marriage}
% \caption{Ratio of boys in ``first'' births}
% \label{fig:sexRatioMarriage}
% \end{figure}

@@

One of the main questions here is how fertility affects sex selection use.
I start by showing the relationship between self-reported desired fertility and sex 
selection, which also illustrates how to interpret the results.
Although self-reported desired fertility is an imperfect proxy for actual
fertility behavior, the analysis of desired fertility indicates that there is 
a relationship between fertility decisions and the use of sex selection and motivates 
examining the relationship more closely.

Because of the desired fertility measure's potential shortcomings, I next examine the 
relationship between actual fertility and use of sex selection for better-educated women.
There are two steps in this process.
First, I show how fertility has declined for better-educated women with one son---a group 
unlikely to use sex selection, which means that sex selection will not affect the 
progression rates to next birth for these women.
Second, I show how, corresponding to the fertility decline, there is an increase in
sex-selective abortions for better-educated women without a son.
I next show how women with lower cost of children appear to follow a high fertility, low
use of sex selection strategy compared to the low fertility, high use of sex selection of
the better-educated women.
The results all show clear evidence of son preference in fertility and sex selection use, 
but exactly what form son preference takes is important, so I next discuss what we can
learn about son preference from the sex selection behavior.
Finally, I predict lifetime fertility, number of abortions, and final sex ratios for
the better-educated women.

Except for the predictions, results are presented using graphs of estimated percentage boys 
born by quarter, the associated 95 percent confidence interval, and the survival curve for 
a ``representative woman'' using the method detailed in Section \ref{sec:strategy}.
The ``representative woman'' characteristics are based on means of continuous 
explanatory variables and the majority category for categorical explanatory variables.
Common to all ``representative women'' is that they do not own land and belong to neither 
a scheduled caste nor a scheduled tribe.
The graphs also show the expected natural rate of boys, approximately 51.2 percent, for
comparison.
Graphs for all groups and spells, even if not discussed here, are available in 
the online Appendix.

\subsection{One Son or Many Sons?}

The extent to which people use sex selection for different sex compositions
of previous children is of interest for two reasons.
First and foremost, it can help us understand what form son preference takes in India.
This is important in itself, but is especially important as we try to predict 
what will happen to sex selection use in the future.
[need an explanation of this based on what I have in the introduction]
Second,
if families use sex selection to reach two or more sons, progression rates for families 
with one son may overestimate the decline in fertility.

Women with 8 or more years of education is the group the most likely to use sex 
selection if they have no son as shown above.
Hence, if sex selection is also used to achieve more than one son, we should be
most likely to find evidence for this group of women as well.
Figure \ref{fig:boys_latest} therefore shows sex ratios and survival curves for 
the 1995--2006 period for urban and rural women with 8 or more years of education, 
conditional on having at least one son.
The first column shows outcomes for the second spell if the first born is a son,
the second column shows third spell when the first two children were a son and 
a daughter, and the third column shows third spell when the first two children were sons.

% \begin{figure}[htpb]
% \centering
% \caption*{Urban}
% \setcounter{subfigure}{-2}
% \subfloat[Spell 2 - 1 boy (N=3,969)]{
%     \begin{minipage}{0.30\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell2_g3_high_r2_pc}}\\
%         \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell2_g3_high_r2_s}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \setcounter{subfigure}{-1}
% \subfloat[Spell 3 - 1 boy/1 girl (N=2,357)]{
%     \begin{minipage}{0.30\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell3_g3_high_r4_pc}}\\
%         \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_r4_s}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \setcounter{subfigure}{0}
% \subfloat[Spell 3 - 2 boys (N=1,098)]{
%     \begin{minipage}{0.30\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell3_g3_high_r2_pc}}\\
%         \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_r2_s}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \caption*{Rural}
% \setcounter{subfigure}{1}
% \subfloat[Spell 2 - 1 boy (N=3,044)]{
%     \begin{minipage}{0.30\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell2_g3_high_r1_pc}}\\
%         \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell2_g3_high_r1_s}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \setcounter{subfigure}{2}
% \subfloat[Spell 3 - 1 boy/1 girl (N=1,819)]{
%     \begin{minipage}{0.30\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell3_g3_high_r3_pc}}\\
%         \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_r3_s}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \setcounter{subfigure}{3}
% \subfloat[Spell 3 - 2 boys (N=790)]{
%     \begin{minipage}{0.30\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell3_g3_high_r1_pc}}\\
%         \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_r1_s}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \caption{Predicted probability of having a boy and probability of
% no birth by quarter (3 month period) for women with 8 or more years
% of education for 1995--2006. 
% N is number of women in the relevant group in the underlying samples.
% }
% \label{fig:boys_latest}
% \end{figure}

For the second spell the predicted sex ratios closely follow the natural sex 
ratio for both urban and rural women.
Furthermore, the survival curves follow the expected pattern if no sex selection 
is taking place.
There is also no evidence of sex selection for the third spell for women with two sons;
the sex ratios follow the natural sex ratio for both urban and rural women.
Sample sizes are, however, less than \sfrac{1}{3} of the sample sizes for the second 
spell and the progression rate to a third birth is low at only 30\% for urban women and 
55\% for rural women, resulting in wide confidence intervals.
The survival curves are particularly flat for the first year, meaning that the apparent 
deviations from the natural sex ratio in the beginning of the spell are based on very few births.

% [Spell 3 - 1b/1g most interesting]

Even though there is no evidence of using sex selection to have an additional son when the 
previous children are all sons, it is possible that families want more sons than daughters 
and therefore use sex selection for the third spell when they already have one of each.
Urban women show little evidence of this being the case.
Although the sex ratio is above the natural from quarter 11 until the end of the spell,
only about \sfrac{1}{3} of the births occur in this interval.
Furthermore, the dip below the natural sex ratio prior to this point also accounts for
approximately \sfrac{1}{3} of the births.
Hence, these two parts cancel each other out.
With the first part being to the natural sex ratio the end result is that there
is little evidence for urban women using sex selection if they already have a son and
a daughter.

For rural women, however, the dip and the associated number of births is not
large enough to bring the overall sex ratio for the spell back to natural sex ratio.
The sex ratio for the entire spell is therefore likely more male dominated
than the natural rate, although it is only statistically significantly higher for two 
individual quarters of the spell.
There are two possible explanations for rural women appearing more likely than urban
women to use sex selection on the third spell if they already have a son and a daughter.
First, rural women could have a stronger son preference than urban women and want two sons 
rather than only one.
Second, mortality is higher in rural than urban areas and this higher mortality, either 
experienced or anticipated, make rural women use sex selection to secure a second son, 
essentially as an insurance.
Separating these two possible explanations is difficult, but the ``heir and a spare''
explanation appears  more likely than difference in preferences given that urban women use 
sex selection more than rural women for the second birth in the absence of a son.
An alternative explanation is that what we are observing is simply random variation.
This would explain why we get see differences in sex ratios between urban and rural
women with one son across the third and fourth spells as shown in the online Appendix.%
\footnote{
Urban women with a son and two daughters as their first 3 births show a sex ratio
that is significantly higher than the natural, but no evidence of sex
selection if they have two sons and a daughter as their first children.
Rural women show only a marginally higher sex ratio with a son and two daughters and no 
effect at all with two sons and a daughter.
These results are, however, based on very small samples.
There are 590 urban women with a son and two daughters, of which less than 40\% have a
fourth child, 344 urban women with two sons and one daughter, and only 20\% of those
have a fourth child.
For rural women the numbers are 564, with 40\% having a fourth child, and 367 with
approximately 30\% having a fourth birth.
This comes to approximately 300 urban births and only slightly more for rural across
the two sex compositions.
}

Son preference can manifest itself through other channels than sex selection.
Differential investments in education and health across boys and girls, resulting in
lower education and higher mortality of girls relatively to boys, could be thought 
of as evidence of son preference, although \cite{rosenzweig1982a} argue that both may 
represent parents' rational response to market opportunities rather than inherent 
preferences of parents.
Whatever way son preference presents itself, it appears that the dominant  
preference---as expressed through use of sex selection---is for one son, rather 
than multiple sons.

@@

@@

@@

@@

If women have preferences over both the number of children they have \emph{and} the sex 
composition of these children, they face a trade-off between the cost---both monetary 
and psychological---of sex selection and the cost of children.
This paper introduces a novel approach to understanding the relationship between
fertility and use of sex selection, and how sex selection interact with birth spacing.
The proposed method allows for joint estimation of fertility and sex-selective abortions 
using a non-proportional, competing risk hazard model.

Three results stand out.
First, lower fertility is an important factor in the decision to use sex-selective abortions.
% Women wanting more than two births did not use sex selection on the second birth, while
% those who want two or fewer use sex selection intensively.
In both urban and rural areas, only women with the lowest predicted fertility---those
with 8 or more years of education---use sex selection, and as fertility falls they use it 
for earlier parities.
Women with less education instead follow a high fertility strategy to ensure they have
a son and do not appear to use sex selection.
Previous research was unable to explain why sex selection in India only occurred 
among higher education women because it failed to tie the use of sex selection and the 
fertility decision together.

Second, the legal steps taken to combat sex-selective abortion have not been able
to reverse the practice.
The use of sex selection has been increasing over time and is higher now than before the
PNDT Act was passed.
The predicted number of sex-selective abortions per 100 women during their childbearing 
years is now around 10 for women with 8 or more years of education in both urban and
rural areas.
This is especially interesting given that the cost of prenatal sex determination 
has likely increased in response to the PNDT Act.
This also emphasizes why accounting for the possibility that parents may change
their decision to use sex selection within a spell is important. 
As the theory shows an increase in the cost of prenatal sex determination 
increases the likelihood of parents changing their decision to use sex selection.%
\footnote{
A feature of the new approach is that it allows for changes in the decision 
to use sex selection within a spell.
This is most clearly seen for better-educated urban women whose only child is a girl, 
for whom the use of sex selection declines as the second spell becomes longer.
}

Third,  there is little evidence that women with one or more sons use sex selection,
and their probability of another birth declines substantially once they have a son.
The main exception is for rural women with one son and one daughter, presumably as an 
insurance against mortality.
Hence, parents appear to have a preference for one son rather than multiple sons.

These results imply that the way we have been measuring son preference may not be useful
for understanding how people make decisions on use of sex selection.
Recent research suggests that son preference in India, when measured as ideally having 
more boys than girls, is decreasing over time and with higher education \citep{bhat03,pande07}.%
\footnote{
This measure of son preference is commonly used in the literature. 
See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
In NFHS, and Demographic and Health Surveys in general, son preference is measured as part 
of the general fertility questions.
The first question is how many children a woman would have if she could choose exactly how 
many.
This if followed by the sex preference question:
``How many of these children would you like to be boys, how many would you like to be girls 
and for how many would the sex not matter?''
}
Despite this, use of sex selection has increased and for exactly the group 
argued to show declining preference for sons.
The culprit may be that parents want \emph{one} son and will use sex selection 
intensively to get that son, but refrain from using sex selection to have more sons, 
and our standard measure of son preference cannot captured that.
A better approach may be to directly ask if there is a minimum number of each sex that 
respondents want, possibly in combination with ideal number questions.

This is of more than theoretical interest.
If son preference is expressed as a very strong desire for having at least one
son then reductions in fertility will lead to a further increase in the use of sex 
selection across all groups that hold that preference.
What is more, as shown in the theory section, it is possible to get increasingly
unequal sex ratio when fertility falls even if parents do not want more sons
than daughters as long as they only use abortion on female fetuses.
A deeper understanding of exactly what type of son preference is responsible 
for the increase use of sex selection is clearly an important question for
future data collection and research.

An often-repeated, but poorly substantiated, argument for why there is sex selection 
in India is the dowry system.
This explanation fits poorly with observed behavior.
If true, dowries increase the cost of having girls relative to boys because you have to 
pay to ensure your girls are married.
This implies that poorer households should be the most likely to use sex selection, and 
that the use of sex selection should increase the more girls they have.%
\footnote{
The exception would be if dowries increase more than proportionally with education level 
and income, but there does not appear to be evidence of this.
} 
Similarly, for all households there would be a financial incentive to achieve a higher 
number of sons than daughters.
As shown here, however, sex selection is predominantly used by better educated and 
therefore, on average, wealthier households and only to secure one son rather than multiple 
sons.
Finally, fertility's rapid decline after the introduction of prenatal sex 
determination, combined with low cost of sex selection, suggests that parents'
perceived costs of boys and girls are similar.%
\footnote{
See the discussion in Section \ref{sec:changes} for more detail.
}
Hence, banning dowries, as India has done, is unlikely to affect the use of sex 
selection.

The results presented here lead to a number of important questions for future research.
One is whether marriage market sex ratios affect the use of sex selection.
Parents may care not only about the number and type of children they have but 
also whether they will have grandchildren.
If they care about grandchildren, then a boy could still be preferred over a girl, but a 
married girl would be preferred over an unmarried boy \citep{Bhaskar2011}.
The implication is that parents respond to changes in the expected sex ratio of their
children's marriage market, even given their preference for a son.
This hypothesis can be tested using measures of the observed distributions of
boys and girls and the method proposed here.

A second question for future research, which is especially important when designing 
policies aimed at reducing the use of sex selection, is who the agents are in the decisions 
to use sex selection.
Here, I have treated the household as a single unit, but it is possible, and indeed likely, 
that husband and wife have different preferences over both the factors that determine  
use of sex selection and sex selection itself.
Some of the popular press have assumed that the decision to use sex selection rests 
with the husband.
This, however, does not fit well with the general pattern that women, on average, prefer to 
have fewer children than men do and that declining fertility appear to be the main
driver of sex selection.
A better understanding of who holds what preferences in the family would be a good 
first step to disentangling who makes decisions about sex selection.

A third question is whether there are interactions between sex selection and child outcomes,
such a survival and education.
It has been argued that access to sex selection may benefit girls because those girls
who are born are more likely to be wanted by their parents.
For mortality it is, however, possible that improvements are the result of a purely 
mechanical effect.
As I have shown, increased use of sex selection lengthens the birth spacing after a girl 
and that by itself can improve her survival chances, even if she is no more wanted than before.
The method suggested here can predict the likelihood of a woman using prenatal sex 
determination for a given birth based on her characteristics.
This predicted probability can then be used in estimations of the determinants
of child health to directly test whether increasing use of sex selection is 
beneficial for the girls who are born.
It is unlikely, however, that a substantial mortality effect exists since sex 
selection is mainly used by well-educated women, who tend to have lower mortality, 
but there might be effects on education investments.

% [developments in the future]
In conclusion, the results provide strong clues to how sex selection use will change in the future.
Because lower fertility is responsible for the increase in sex-selective abortions, 
it is likely that we will see further increases in the practice as more families 
want fewer children, either because of urbanization or because of increases in female 
education.
With already low fertility for better-educated urban women, a substantial future increase 
in sex-selective abortions per woman is unlikely, 
but a higher proportion of women will belong to this group in the future.
For rural, better-educated women fertility is still falling.
To the extent that it falls to the same level as for urban
women, I expect a corresponding increase in sex selection.
Finally, and most importantly, we are beginning to see evidence of lower fertility 
for women with lower levels of education.
If women with less education hold the same preference for one son as the better-educated,
once the less-educated women's fertility begins to drop to three---and even two---we are 
likely to see a substantially increased use of sex selection in India.

\newpage
\onehalfspacing
\bibliographystyle{aer}
\bibliography{collection}

@@

>>>>

\title{Son Preference, Birth Spacing, and Sex-Selective Abortions%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of Aarhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support from the University of Washington Royalty Research Fund and the 
Development Research Group of the World Bank is gratefully acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
% Prior versions of this paper were presented under the title ``The 
% Determinants of sex-selective Abortions.''
}
}

@@

\author{Claus C P\"ortner\\
    Department of Economics\\
    Albers School of Business and Economics\\
    Seattle University, P.O. Box 222000\\
    Seattle, WA 98122\\
    \href{mailto:claus@clausportner.com}{\texttt{claus@clausportner.com}}\\
    \href{http://www.clausportner.com}{\texttt{www.clausportner.com}}\\
    \& \\
    Center for Studies in Demography and Ecology \\
    University of Washington\\ \vspace{2cm}
    }

\date{September 2017\\
\bigskip
Preliminary}

@@

% Demography abstract
\noindent This paper examines how birth spacing interact with the use of 
sex-selective abortions. 
I introduce a statistical method that incorporates that sex-selective abortions 
affect both the likelihood of a son and spacing between births.
Using India's National Family and Health Surveys,
I show that use of sex selection leads to \emph{longer} spacing after a daughter 
is born than after a son, but for families that do not use sex selection
duration to next birth is still shorter when no boys are present.
Women with 8 or more years of education, both in urban and rural areas, are 
the main users of sex-selective abortions, whereas women with less education 
do not appear to use sex selection.

@@

[Outline of findings here]

% [EXPLAIN RESULTS HERE]
% Main findings!
% There are three main findings.
% 
% There is, however, still no evidence of sex selection being used on the first birth.
% 
% Secondly, sex selection appears to be used for securing one son, rather than a large 
% number of sons.
% There is only limited use of sex-selective abortions for better-educated
% women with one or more sons.
% The exception is rural women with one son and one daughter, presumably to compensate for 
% the higher child mortality in rural areas.
% These results are in line with the differential stopping behavior observed in many studies
% before sex selection became available \citep{repetto72,arnold98,dreze01}.
%

@@

[The problem with comparing survival curves]

@@

One solution to this problem is to condition on the parity progression
rates.
That is, instead of comparing say how long it takes for 50\% of women
who began the spell to have the next birth across boy and girl composition, 
which would work well if the next birth was nearly universal, we 
could calculate how long it takes for 50\% of those women who will 
eventually have a next child to have that child based on the estimated
survival curves.
This number can then be compared across sex compositions to 
understand if the spell length for girl composition increases
more than for boy composition with the introduction of sex selection.
Obviously, this should be calculated for different percentages
and the easiest way to interpret the result is to graph survival
conditional on parity progression occurring.
All compositions therefore begin at 100\% and end at 0\% because
we condition on having the next child.

@@

\begin{figure}[htp]%
\centering
\includegraphics[width=0.9\textwidth]{sex_ratio_marriage}
\caption{Ratio of boys in ``first'' births}
\label{fig:sexRatioMarriage}
\end{figure}

@@

[Only a subset of the planned analyses to demonstrate the method and interpretation]

@@

Figure \ref{fig:results_spell2_high_urban} 
shows the predicted probabilities of having a boy and standard survival curves
for the three periods for urban women with 8 or more years of education.%
\footnote{
The corresponding graphs for rural women and for women with no education are
shown in Figures \ref{fig:results_spell2_high_rural}, \ref{fig:results_spell2_low_urban},
and \ref{fig:results_spell2_low_rural}
}
Two main results stand out.
First, women whose first child was a boy do not appear to use sex selection during
the second spell.
The predicted percent boys is consistently close to the natural level in all three
panels (\ref{fig:results_spell2_high_urban}(d)-(f)).
Second, as prenatal sex determination became available there is more and more
evidence of sex selection affecting the sex ratio for women with a girl as their
first child.
In panel (a), which covers the period before sex selection became generally
available, there is no significant evidence of a sex ratio biased towards males.
Panel (b) shows a more male biased sex ratio but only a few of quarters show
a statistically significant difference from the natural rate and the shallow 
slope of the survival curve indicates that relatively few births took place
during these quarters.
Panel (c), however, shows clear evidence of the effect of access to prenatal
sex determination with a male biased sex ratio for more of the two first years.%
\footnote{
Recall that quarter 0 is equivalent to 9 months (3 quarters) after the birth
of the first child.
}

@@

Figures \ref{fig:results_spell2_low_pps} and \ref{fig:results_spell2_high_pps}
show survival curves for the second spell (from first birth) for women with
no education and women with 8 or more years of education.
To allow for comparison these survival curves show the ratio of women who
have not had a birth yet, given that they will eventually are predicted to
have a birth.
Women with no education show relatively similar survival curves whether their
first child was a girl or a boy, but to the extent that a pattern is present
it shows, as expected, shorter spacing a girl than a boy.
Rural women with 8 or more years of education show essentially the same 
pattern.
The lack of clear results is not surprising given that all three groups of
women are likely to have more than two children.

The effect of sex selection on spacing first show up for urban women with
8 or more years of education.
Panel (a) in Figure \ref{fig:results_spell2_high_pps} show shorter spacing
for the second spell when the first child is a girl than a boy, and this
pattern is also present in Panel (b).
Panel (c) on the other hand show first evidence of \emph{longer} spacing 
after the birth of a girl than after a boy, consistent with the male-biased
sex ratio in Panel (c) in Figure \ref{fig:fig:results_spell2_high_urban}.

@@

% 3rd spell

\begin{figure}[htpb]
\centering
\caption*{Urban}
\setcounter{subfigure}{-1}
\subfloat[1972--1984]{
    \begin{minipage}{0.31\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob.\ no birth yet]{\includegraphics[width=\textwidth]{spell3_g1_low_pps_urban}} 
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
} 
\setcounter{subfigure}{-0}
\subfloat[1985--1994]{
    \begin{minipage}{0.31\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g2_low_pps_urban}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\setcounter{subfigure}{1}
\subfloat[1995--2006]{
    \begin{minipage}{0.31\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_low_pps_urban}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\caption*{Rural}
\setcounter{subfigure}{2}
\subfloat[1972--1984]{
    \begin{minipage}{0.31\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g1_low_pps_rural}} 
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
} 
\setcounter{subfigure}{3}
\subfloat[1985--1994]{
    \begin{minipage}{0.31\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g2_low_pps_rural}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\setcounter{subfigure}{4}
\subfloat[1995--2006]{
    \begin{minipage}{0.31\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_low_pps_rural}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\caption{Survival curves conditional on parity progression
for women with no education by quarter (3 month period).
Left column shows results prior to sex selection available, middle column before
sex selection illegal and right column after sex selection illegal.
N indicates the number of women in the relevant group in the underlying samples.
}
\label{fig:results_spell3_low_pps}
\end{figure}

\begin{figure}[htpb]
\centering
\caption*{Urban}
\setcounter{subfigure}{-1}
\subfloat[1972--1984]{
    \begin{minipage}{0.31\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob.\ no birth yet]{\includegraphics[width=\textwidth]{spell3_g1_high_pps_urban}} 
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
} 
\setcounter{subfigure}{-0}
\subfloat[1985--1994]{
    \begin{minipage}{0.31\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g2_high_pps_urban}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\setcounter{subfigure}{1}
\subfloat[1995--2006]{
    \begin{minipage}{0.31\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_pps_urban}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\caption*{Rural}
\setcounter{subfigure}{2}
\subfloat[1972--1984]{
    \begin{minipage}{0.31\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g1_high_pps_rural}} 
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
} 
\setcounter{subfigure}{3}
\subfloat[1985--1994]{
    \begin{minipage}{0.31\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g2_high_pps_rural}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\setcounter{subfigure}{4}
\subfloat[1995--2006]{
    \begin{minipage}{0.31\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_pps_rural}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\caption{Survival curves conditional on parity progression
for women with 8 or more years of education by quarter (3 month period).
Left column shows results prior to sex selection available, middle column before
sex selection illegal and right column after sex selection illegal.
N indicates the number of women in the relevant group in the underlying samples.
}
\label{fig:results_spell3_high_pps}
\end{figure}

The effect of sex selection on spacing is even more pronounced in
Figure \ref{fig:results_spell3_high_pps}, which shows the conditional
survival curves for the third spell (from second birth) for women with
8 or more years of education.
Before prenatal sex determination became available there is little difference
in spacing pattern no matter the sex composition of the first two children.
This changes after the introduction as shown in Panel (b), where the spacing
is longer when the first two children are girls than either one boy/one girl
or two boys.
Even more substantial are the results from Panel (c), which shows much 
longer spacing when the first two children are girls.
These results are consistent with the biased sex ratios shown in 
Figure \ref{fig:results_spell3_high_urban}.

\clearpage

@@

[To be added]

\clearpage

\onehalfspacing
\bibliographystyle{aer}
\bibliography{collection}

@@

\clearpage

\subsection{Second Spell}
